---
title: "Missing Values"
author: "Pritpal Singh"
date: "2023-9-19"
note: "open this file in RStudio"
---

If we encountered unusual values in our dataset, and simply want to move to the rest of the analysis, we have two options.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

-   Drop the entire row with the strange values

    ```{r}
    diamonds2 <- diamonds %>%
      filter(between(y, 3, 20))

    diamonds2
    ```

Well this option is not recommended!!!!

-   Instead, replace the unusual values with missing values. The easiest way to do this is to use `mutate()` to replace the variables with a modified copy. We can us the `ifelse()` function to replace unusual values with NA

    ```{r}
    diamonds2 <- diamonds %>%
      mutate(y = ifelse(y < 3| y > 20, NA, y))

    diamonds2
    ```

`ifelse()` has three arguments. The first argument `test` should be a logical vector. The result will contain the value of the second argument, `yes`, when `test` is `TRUE`, and the value of the third argument, `no`, when it is `FALSE`.

Like R, **ggplot2** subscribes to the philosophy that missing values should never silently go missing. It\'s not obvious where you should plot missing values, so **ggplot2** doesn\'t include them in the plot, but it does warn that they\'ve been removed.

```{r}
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) +
  geom_point()
```

To suppress that warning, set `na.rm = TRUE`

```{r}
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) +
  geom_point(na.rm = TRUE)
```

Other times we want to understand what makes observations with missing values different from observations with recorded values. For example, in `nycflights13 :: flights`, missing values in the `dep_time` variable indicate that the flight was cancelled. So we might want to compare the scheduled departure times for cancelled and noncancelled times. We can do this by making a new variable `is.na()`.

```{r}
nycflights13::flights %>%
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>%
  ggplot(mapping = aes(sched_dep_time)) + 
  geom_freqpoly(
    mapping = aes(color = cancelled),
    binwidth = 1/4
  )
```

However, this plot is not great because there are many more non-cancelled flights than the cancelled flights.
