---
title: "Pattern and Models"
author: "Pritpal Singh"
date: "2023-9-22"
note: open this file in RStudio
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

Patterns in our data provide clues about the relationships. If a systematic relationship exists between two variables it will appear as a pattern in the data. If we spot a pattern, ask yourself\\

-   Could this pattern be due to coincidence (i.s random chance)?

-   How can we describe the relationship implied by the pattern?

-   What other variables might effect the relationship?

-   Does the relationship change if we look at individual subgroups of the data?

A scatterplot of Old Faithful eruption lengths versus the wait time between eruption shows a pattern.: longer eruptions. The scatterplot also displays the two clusters that we noticed earlier.

```{r}
?faithful
```

```{r}
ggplot(data = faithful) +
  geom_point(mapping = aes(x = eruptions, y = waiting))
```

Patterns provides one of the most useful tools for data scientists because they reveal covariation. If we think of variation as phenomenon that creates uncertainty, covariation is a phenomenon that reduces it. Of two variables covary, we can use the values of one variable to make better predictions about the values of the second. If the covariation is due to a casual relationship, then we can use the value of one variable to control the value of the second variable.

Models are a tool for extracting patterns out of data. For example, consider the diamonds data. It's hard to understand the relationship between cut and price., because the `cut` and `carat`, and carat and price, are tightly related. it's possible to use a model to remove the very strong relationship between price and carat so we can explore the subtleties that remain. The following code fits a model the predicts `price` from `carat` and then computes the residuals (the difference between the predicted value and the actual value). The residuals give us a view of the price of the diamonds, once the effect of carat has been removed.

```{r}
library(modelr)

mod <- lm(log(price) ~ log(carat), data = diamonds)

diamonds2 <- diamonds %>%
  add_residuals(mod) %>%
  mutate(resid = exp(resid))

ggplot(data = diamonds2) + 
  geom_point(mapping = aes(x = carat, y = resid))
```

Once we have removed the strong relationship between carat and price, we can see what we expect in the relationship between `cut` and `price` -- relative to their size, better quality diamonds are more expensive.

```{r}
ggplot(data = diamonds2) + 
  geom_boxplot(mapping = aes(x = cut, y = resid))
```
