---
title: "Select Coumns with Select()"
author: "Pritpal Singh"
date: "2023-9-10"
note: "open this file in RStudio"
---

`select()` allows us to rapidly zoom in on a useful subset using operations based on the names of the variables.

```{r}
library(tidyverse)
library(nycflights13)
```

`select()` is not terribly useful with the flight data because we only have 19 variables, but we can still get the general idea.

```{r}
# select columns by name
select(flights, year, month, day)
```

```{r}
# select al columns between year and day (inclusive)
select(flights, year:day)
```

```{r}
# Select all columns except those from year to day(inclusive)
select(flights, -(year:day))
```

There are a number of helper functions we can use within `select()`

-   `starts_with("abc")` matches names begin with "abc"

-   `end_with("xyz")` matched names that end with "xyz"

-   `contains("ijk")` matches names that contain "ijk"

-   `matches("(.)\\1")` selects variables that match a regular expression. This one matches any variables that contains repeated characters.

-   `num_range("x", 1:3)` matches `x1`, `x2` and `x3`.

`select()` can be used to rename variables, but it's rarely useful because it drops all of the variables not explicitly mentioned. Instead, use `rename()`, which is a variant of `select()` that keeps all the variables that are not explicitly mentioned.

```{r}
rename(flights, tail_num = tailnum)
```

Another option is to use `select()` in conjunction with the `everything()` helper. This is useful if you have a handful of variables we'd like to move to the start of the data frame.

```{r}
select(flights, time_hour, air_time, everything())
```

#### TODO:

**Please complete the exercise on your own**

## Exercise

1.  Brainstorm as many ways as possible to select `dep_time, dep_delay, arr_time,` and `arr_delay` from `flights`.

2.  What happens if you include the name of a variable multiple times in `select()` call?

3.  What does the `one_of()` function do? Why might it be helpful in conjunction with this vector?

    `vars <- c(`

    `"year", "month", "day", "dep_delay_, "arr_delay"`

    `)`

4.  Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?

    `select(flights, contains("TIME"))`

## Answer

**1**

> Brainstorm as many ways as possible to select `dep_time, dep_delay, arr_time,` and `arr_delay` from `flights`.

```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
```

```{r}
select(flights, "dep_time", "dep_delay", "arr_time", "arr_delay")
```

```{r}
select(flights, 4, 6, 7, 9)
```

```{r}
select(flights, all_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
```

```{r}
select(flights, any_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
```

```{r}
variables <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
select(flights, all_of(variables))
```

```{r}
select(flights, starts_with("dep_"), starts_with("arr_"))
```

```{r}
select(flights, matches("^(dep|arr)_(time|delay)$"))
```

```{r}
variables <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
select(flights, !!variables)
```

```{r}
variables <- syms(c("dep_time", "dep_delay", "arr_time", "arr_delay"))
select(flights, !!!variables)
```

```{r}
select(flights, ends_with("arr_time"), ends_with("dep_time"))
```

```{r}
select(flights, contains("_time"), contains("arr_"))
```

**2**

> What happens if you include the name of a variable multiple times in a `select()` call?

```{r}
select(flights, year, month, day, year, year)
```

The `select()` call ignores the duplication. Any duplicated variables are only included once, in the first location they appear. The `select()` function does not raise an error or warning or print any message if there are duplicated variables.

This behavior is useful because it means that we can use `select()` with `everything()` in order to easily change the order of columns without having to specify the names of all the columns.

```{r}
select(flights, arr_delay, everything())
```

**3**

> What does the `one_of()` function do? Why might it be helpful in conjunction with this vector?

The `one_of()` function selects variables with a character vector rather than unquoted variable name arguments. This function is useful because it is easier to programmatically generate character vectors with variable names than to generate unquoted variable names, which are easier to type.

```{r}
#vars <- c("year", "month", "day", "dep_delay", "arr_delay")
#select(flights, one_of(vars))
```

In the most recent versions of **dplyr**, `one_of` has been deprecated in favor of two functions: `all_of()` and `any_of()`. These functions behave similarly if all variables are present in the data frame.

```{r}
#select(flights, any_of(vars))
```

```{r}
#select(flights, all_of(vars))
```

These functions differ in their strictness. The function `all_of()` will raise an error if one of the variable names is not present, while `any_of()` will ignore it.

```{r}
#vars2 <- c("year", "month", "day", "variable_not_in_the_dataframe")
#select(flights, all_of(vars2))
```

Uncomment the above code to see the output

```{r}
#select(flights, any_of(vars2))
```

The deprecated function `one_of()` will raise a warning if an unknown column is encountered.

```{r}
#select(flights, one_of(vars2))
```

Uncomment the above code to see the output

In the most recent versions of **dplyr**, the `one_of()` function is less necessary due to new behavior in the selection functions. The `select()` function can now accept the name of a vector containing the variable names you wish to select:

```{r}
#select(flights, vars)
```

Uncomment the above code to see the output

However there is a problem with the previous code. The name `vars` could refer to a column named `vars` in `flights` or a different variable named `vars`. What th code does will depend on whether or not `vars` is a column in `flights`. If `vars` was a column in `flights`, then that code would only select the `vars` column. For example:

```{r}
#flights <- mutate(flights, vars = 1)
#select(flights, vars)
```

However, `vars` is not a column in `flights`, as is the case, then `select` will use the value the value of the , and select those columns. If it has the same name or to ensure that it will not conflict with the names of the columns in the data frame, use the `!!!` (bang-bang-bang) operator.

```{r}
#select(flights, !!!vars)
```

**4**

> Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?

```{r}
select(flights, contains("TIME"))
```

The default behavior for `contains()` is to ignore case. This may or may not surprise you. If this behavior does not surprise you, that could be why it is the default. Users searching for variable names probably have a better sense of the letters in the variable than their capitalization. A second, technical, reason is that dplyr works with more than R data frames. It can also work with a variety of [databases](https://db.rstudio.com/dplyr/). Some of these database engines have case insensitive column names, so making functions that match variable names case insensitive by default will make the behavior of `select()` consistent regardless of whether the table is stored as an R data frame or in a database.

To change the behavior add the argument `ignore.case = FALSE`.

```{r}
select(flights, contains("TIME", ignore.case = FALSE))
```
