---
title: "Coordinate Systems"
author: "Pritpal Singh"
date: "2023-09-09"
note: open this file in RStudio
---

```{r}
library(ggplot2)
```

Coordinate Systems are probably the most complicated part of **ggplot2**. The default coordinate system is the Cartesian Coordinate system where the x and y position act independently to find the location of each point. There are number of other coordinate systems that are occasionally helpful.

1.  `coord_flip()` switches the x- and y-axes. This is useful if we want horizontal boxplots. It's also useful for long labels -- it's hard to get them to fit without overlapping on the x-axis.

    ```{r}
    ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
      geom_boxplot()

    ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
      geom_boxplot() + 
      coord_flip()
    ```

2.  `coord_quickmap()` sets the aspect ratio correctly for maps. This is very important if we are plotting spatial data with **ggplot2**.

    ```{r}
    nz <- map_data("nz")

    ggplot(nz, aes(long, lat, group = group)) +
     geom_polygon(fill = "white", color = "black")

    ggplot(nz, aes(long, lat, group = group)) +
     geom_polygon(fill = "white", color = "black") +
     coord_quickmap()
    ```

3.  `coord_polar()` uses polar coordinates. Polar coordinates reveal an interesting connection between a bar chart and a Coxcomb chart.

    ```{r}
    bar <- ggplot(data = diamonds) + 
      geom_bar(
        mapping = aes(x = cut, fill = cut),
        show.legend = FALSE,
        width = 1
      ) + 
      theme(aspect.ratio = 1) + 
      labs(x = NULL, y = NULL)

    bar + coord_flip()
    bar + coord_polar()
    ```

## Exercises

1.  Turn a stacked bar chart into a pie chart using `coord_polar()`.

2.  What does `labs()` do? Read the documentation.

3.  What's the difference between `coord_quickmap()` and `coord_map()`?

4.  What does the following plot tell you about the relationship between city and highway mpg? Why is `coord_fixed()` important? What does `geom_abline()` do?

    ```{r}
    ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
      geom_point() + 
      geom_abline() + 
      coord_fixed()
    ```

## Answer

**1**

> Turn a stacked bar chart into a pie chart using `coord_polar()`.

A pie chart is a stacked bar chart with the addition of polar coordinated, For example:

```{r}
ggplot(data = mpg, mapping = aes(x = factor(1), fill = drv)) +
  geom_bar()
```

Now add `coord_polar(theta = 'y')` to create pie chart

```{r}
ggplot(data = mpg, mapping = aes(x = factor(1), fill = drv)) +
  geom_bar() + 
  coord_polar(theta = 'y')
```

The argument `theta = 'y'` maps `y` to the angle of each section. If `coord_polar()` is specified without `theta = 'y'`, then the resulting plot is called a bulls-eye chart

```{r}
ggplot(data = mpg, mapping = aes(x = factor(1), fill = drv)) +
  geom_bar(width = 1) + 
  coord_polar()
```

**2**

> What does `labs()` do? Read the documentation.

The `labs()` function adds axis titles, plot titles, and a caption to the plot.

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() + 
  coord_flip() + 
  labs(y = 'Highway MPG',
       x = "Class",
       title = "Highway MPG by car class",
       subtitle = '199-2008',
       caption = "Source: http://fueleconomy.gov"
       )
```

The arguments to `labs()` are optional, so we can add as many or as few of these as are needed

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() + 
  coord_flip() + 
  labs(y = 'Highway MPG',
       x = "Class",
       title = "Highway MPG by car class",
       )
```

**3**

> What's the difference between `coord_quickmap()` and `coord_map()`?

```{r}
?coord_map()
```

```{r}
?coord_quickmap()
```

The `coord_map()` function uses map projections to project the 3D Earth onto a 2D plane. By default, `coord_map()` uses the Mercator projection. This projection is applied to all the geoms in the plot. The `coord_quickmap()` function uses an approximate but faster map projection. This approximation ignores the curvature of Earth and adjusts the map for the latitude/longitude ratio. The `coord_quickmap()` project is faster then `coord_map()` both because the projection is computationally easier, and unlike `coord_map()`, the coordinates of the individual geoms do not need to be transformed.

**4**

> What does the following plot tell you about the relationship between city and highway mpg? Why is `coord_fixed()` important? What does `geom_abline()` do?

```{r}
p <- ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point() + 
  geom_abline() 

p +  coord_fixed()
```

If we did not include `coord_fixed()`, then the line would no longer have an angle of 45 degrees.

```{r}
p
```

On average, humans are best able to perceive difference in angles relative to 45 degrees. The Function \``ggthemes::bank_slopes()` will calculate the optimal aspect ratio to bank slopes to 45 degrees.
